Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Pass on whole text, fixed some comments

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, that is dictionaries (written \nickel{\{field1 = value1, ...,
  fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
with functions (\nickel{fun arg1 ... argn => body}) and
let-definitions (\nickel{let id = value in exp}).\unsure{Arnaud: this
  paragraph looks kind of bad, maybe it'll be better to put each
  syntax entry in an enumeration, we will see if there is still room
  for this}

@@

A probably even more common pattern is a variable which can contain
either a value of type a (say, a number) or the \typescript{null}
value. So much so, in fact, that MyPy defines a type alias
\python{Optional[A]} for \python{Union[A,None]} (\python{None} is how
Python renders the null value).

@@

\unsure{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array<string>} or an \typescript{Array<number>},
what value could we fill \typescript{...} with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{string} or a
\typescript{number}.
That is, when using a union, we have to treat it as if it where any one
of the alternatives, in the case of the example, we should provide
it with a value that is both a \typescript{string} and a \typescript{number}.

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function addElem(list: Array<string> | Array<number>) {
  list.push(...)
}
\end{lstlisting}
\caption{Adding an element to a union of two arrays}
\label{fig:addElem}
\end{figure}

@@

Intersection contracts (and types) are probably less prevalent than
union in practical type systems. However, a function from a union is
equivalent to an intersection. That is \nickel{(A @| B) ->
  C}$\simeq$\nickel{(A -> C) @& (B -> C)}. So in a system with
functions and unions, intersections are already morally present (and,
for that matter, in a system with functions and intersections, unions
are morally present). Some
of our examples in Sections~\ref{sec:issues-sem}
and~\ref{sec:issues-literature} are better expressed in terms of
intersections, so it's best to include them.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal = { species: Str, race: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal @& Pet =
  { species = "Canis Lupus",
    race = "Australian Cattle Dog",
    owner = "Anonymous Author",
    name = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

@@

However appealing union and intersection contracts may be, they happen
to be fundamentally incompatible with the desirable language
features from Section~\ref{sec:feat-lang}. At least in there
full-blown form: in Section~\ref{sec:issues-literature} we will
discuss pragmatic restrictions of union and intersection contracts to
recover some or all of the features.

@@

In Nickel, the failure of a function contract can always be traced back to a
single call. For example, take the function \nickel{f} with a simple contract
attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
error blaming \nickel{f} because the return value of the second call \nickel{f 5}
violates the \nickel{Positive} contract. The first call to \nickel{f} does
not matter, and \nickel{f 5} is a single and independent witness of the contract
violation. The user is pointed to this one location in practice.

@@

In an higher-order intersection contract, blame is raised when:
\begin{description}
    \item[Faulty caller] The argument fails \emph{both} components.
    \item[Faulty implementation] The function fails \emph{one} component that the
        argument previously satisfied.
\end{description}

@@

Curiously, union and intersection type systems, on which
\cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
exhibit this property. Subtyping allows to correctly type check a term like
\nickel{g 1} as both \nickel{Num -> Num} as well as
\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
can be given the type \nickel{Num -> Num @& Bool -> Num}.
\unsure{(Yann) I just reworked the wording, but I'm not totally sold on the
argument}

@@

An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
these nested contracts using the De Morgan law of
Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
intersection as much as possible.  \todo{This assumes the difficulty has been
previously explained} This transformation flattens arbitrarily nested unions and
intersections to a union of intersections, making blame resolution more
manageable.\unsure{(Yann) find better (more precise) wording}

@@

To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. The result is that the reduction rules sometimes feel
ad-hoc, and make us question the composability and extensibility of the system.
In the words of Williams, Morris, and Wadler (see
Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
re-evaluated in each branch of the right hand side of
Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
mechanism.
\resolved{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel,title={Original program}]
let C = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (Str -> Str) @& C
      = fun x => x
in g 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={First component}]
g 0 | Str -> Str
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={Second component}]
g 0 | C
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts-2}
\end{figure}

@@

On the semantic side, one of the so-called the sound monitoring properties
linking the operational semantics to the denotational semantics (supposed to
capture the intuitive interpretation) is at odds with the given interpretation.
% On top of that, a fundamental problem arises on said paper,
% symptom of the difficulty of establishing good enough, and yet
% simple semantics for union and intersection dynamic checks.
% One of their main results, is the fact that their calculus obeys
% all of their expected monitoring properties for higher ordered, union
% and intersection contracts.

@@

Racket is a language based on the Scheme dialect of Lisp. Among established
languages, Racket is probably the one with the most comprehensive contract
system\cite{RacketContracts}. Regarding union and intersections, Racket provides
the \racket{and/c} and \racket{or/c} combinators for contract.  These
combinators correspond to the naive interpretation of unions and intersections
(as in Figure~\ref{fig:naive-impl}): \racket{and/c} (resp.  \racket{or/c})
simply checks sequentially that all (resp. one) contracts pass.  In consequence,
unions or intersections of higher-order contracts don't match intuition, and in
particular intersections don't model overloading. As in
Figure~\ref{fig:naive-impl}, the example given in
Figure~\ref{fig:racket-overloading} always fail because no argument satisfies
both \racket{number?} and \racket{string?}.

@@

When higher-order contracts are combined using \racket{or/c}, Racket imposes
that contracts must distinguishable by their arity. Doing so, there is at most
one candidate that can be selected directly. This is illustrated in
Figure~\ref{code:racket:or/c:working} whose program is accepted. On the other
hand, the program of Figure~\ref{code:racket:or/c:non-working} is rejected.

@@

Racket does provide the \racket{case->} combinator which(partially) supports
overloading. As for the \racket{or/c}, the candidate contracts must have
distinct arities to avoid ambiguity. An example is provided in
Figure~\ref{fig:rafig:racket-case-fun}. The resulting possibilities are similar
to static overloading, where one function can take additional parameters for
example (e.g. as supported for Java methods). Unfortunately, it excludes the
overloading of generic operations with fixed arity such as equality, comparison,
arithmetic operators, and so on.

@@

The implementation of casts relies on abstract interpretation, leading to a
complex specification, whose implementation is straightforward. It is not clear
either how much these casts rely on static type information to function
correctly, possibly hampering their transformation to more general standalone
contracts.

>>>>

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, and its conformed of:

\begin{itemize}
  \item dictionaries, written as:
\begin{lstlisting}[language=nickel]
{field1 = value1, ..., fieldn = valuen}
\end{lstlisting}
  \item arrays:
\begin{lstlisting}[language=nickel]
[x1, x2, ..., xn]
\end{lstlisting}
  \item functions:
\begin{lstlisting}[language=nickel]
fun arg1 ... argn => body
\end{lstlisting}
  \item and let-definitions:
\begin{lstlisting}[language=nickel]
let id = value in exp
\end{lstlisting}  
\end{itemize}

% In this article, we will use the Nickel language~\cite{NickelRepo} as
% illustration and motivation. At its core, Nickel is the JSON data
% model, that is dictionaries (written \nickel{\{field1 = value1, ...,
%   fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
% with functions (\nickel{fun arg1 ... argn => body}) and
% let-definitions (\nickel{let id = value in exp}).
\resolved{Arnaud: this
  paragraph looks kind of bad, maybe it'll be better to put each
  syntax entry in an enumeration, we will see if there is still room
  for this
  Teo: Tried to enumerate the constructs, left the original commented}

@@

A probably even more common pattern is a variable which can contain
either a value of type \typescript{A} (say, a number) or the \typescript{null}
value. So much so, in fact, that MyPy defines a type alias
\python{Optional[A]} for \python{Union[A,None]} (\python{None} is how
Python renders the null value).

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it where any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

\info{Not planned, but nice place to smooth things out towards
intersections}

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function addElem(list: Array< 42 | "Nickel" > | Array< "Nickel" | 3.5 >) {
    list.push((*@ $\square$ @*))
}
\end{lstlisting}
\caption{Adding an element to a union of two arrays}
\label{fig:addElem}
\end{figure}

% \begin{figure}[h]
% \begin{lstlisting}[language=JavaScript]
% function addElem(values: Array<{ Nix: string }> | Array<{ Nickel: number }>) {
%     values.push({Nix: "Packages", Nickel: 32})
% }
% \end{lstlisting}
% \caption{Adding an element to a union of two arrays}
% \label{fig:addElem}
% \end{figure}

@@

Intersection contracts (and types) are probably less prevalent than
union in practical type systems. However, a function from a union is
equivalent to an intersection. That is \nickel{(A @| B) ->
  C} $\simeq$ \nickel{(A -> C) @& (B -> C)}. So in a system with
functions and unions, intersections are already morally present (and,
for that matter, in a system with functions and intersections, unions
are morally present). Some
of our examples in Sections~\ref{sec:issues-sem}
and~\ref{sec:issues-literature} are better expressed in terms of
intersections, so it's best to include them.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal =
  { species: Str, breed: Str, name: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal @& Pet =
  { species = "Canis Lupus",
    breed   = "Australian Cattle Dog",
    owner   = "Anonymous Author",
    name    = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

@@

However appealing union and intersection contracts may be, they happen
to be fundamentally incompatible with the desirable language
features from Section~\ref{sec:feat-lang}. At least in their
full-blown form: in Section~\ref{sec:issues-literature} we will
discuss pragmatic restrictions of union and intersection contracts to
recover some or all of the features.

@@

In \unsure{Maybe specify its Nickel without Union or inter}Nickel,
the failure of a function contract can always be traced back to a
single call. For example, take the function \nickel{f} with a simple contract
attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
error blaming \nickel{f} because the return value of the second call \nickel{f 5}
violates the \nickel{Positive} contract. The first call to \nickel{f} does
not matter, and \nickel{f 5} is a single and independent witness of the contract
violation. The user is pointed to this one location in practice.

@@

In a higher-order intersection contract, blame is raised when:
\begin{description}
    \item[Faulty caller] The argument fails \emph{both} components.
    \item[Faulty implementation] The function fails at least \emph{one}
        component that the argument previously satisfied.
\end{description}

@@

% Curiously, union and intersection type systems, on which
% \cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
% exhibit this property. Subtyping allows to correctly type check a term like
% \nickel{g 1} as both \nickel{Num -> Num} as well as
% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
% can be given the type \nickel{Num -> Num @& Bool -> Num}.
\resolved{(Yann) I just reworked the wording, but I'm not totally sold on the
argument
(Teo) I agree, I think it would need to be expanded for it to actually
say something, I'll remove it}

@@

An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
these nested contracts using the De Morgan law of
Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
intersection as much as possible.
This transformation flattens arbitrarily nested unions and
intersections to a union of intersections, making blame resolution more
manageable.\unsure{(Yann) find better (more precise) wording}

@@

To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. The result is that the reduction rules sometimes feel
ad-hoc, and make us question the composability and extensibility of the system.
In the words of Williams, Morris, and Wadler (see
Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
re-evaluated in each branch of the right hand side of
Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
mechanism.
\resolved{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
\resolved{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel,title={Original program}]
let C = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (Str -> Str) @& C
      = fun x => x
in g 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={First component}]
(g | Str -> Str) 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={Second component}]
(g | C) 0
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts-2}
\end{figure}

@@

On the semantic side, one of the so-called sound monitoring properties
linking the operational semantics to the denotational semantics (supposed to
capture the intuitive interpretation) is at odds with the given interpretation.
% On top of that, a fundamental problem arises on said paper,
% symptom of the difficulty of establishing good enough, and yet
% simple semantics for union and intersection dynamic checks.
% One of their main results, is the fact that their calculus obeys
% all of their expected monitoring properties for higher ordered, union
% and intersection contracts.

@@

Racket is a language based on the Scheme dialect of Lisp. Among established
languages, Racket is probably the one with the most comprehensive contract
system\cite{RacketContracts}. Regarding union and intersections, Racket provides
the \racket{and/c} and \racket{or/c} combinators for contract.  These
combinators correspond to the naive interpretation of unions and intersections
(as in Figure~\ref{fig:naive-impl} for intersections): \racket{and/c} (resp.  \racket{or/c})
simply checks sequentially that all (resp. one) contracts pass.  In consequence,
unions or intersections of higher-order contracts don't match intuition, and in
particular intersections don't model overloading. As in
Figure~\ref{fig:naive-impl}, the example given in
Figure~\ref{fig:racket-overloading} always fail because no argument satisfies
both \racket{number?} and \racket{string?}.

@@

When higher-order contracts are combined using \racket{or/c}, Racket imposes
that contracts must be distinguishable by their arity. Doing so, there is at most
one candidate that can be selected directly. This is illustrated in
Figure~\ref{code:racket:or/c:working} whose program is accepted. On the other
hand, the program of Figure~\ref{code:racket:or/c:non-working} is rejected.

@@

Racket does provide the \racket{case->} combinator which(partially) supports
overloading. As for the \racket{or/c}, the candidate contracts must have
distinct arities to avoid ambiguity. An example is provided in
Figure~\ref{fig:racket-case-fun}. The resulting possibilities are similar
to static overloading, where one function can take additional parameters for
example (e.g. as supported for Java methods). Unfortunately, it excludes the
overloading of generic operations with fixed arity such as equality, comparison,
arithmetic operators, and so on.

@@

The implementation of casts relies on abstract interpretation, leading to a
complex specification, whose implementation is not straightforward. It is not clear
either how much these casts rely on static type information to function
correctly, possibly hampering their transformation to more general standalone
contracts.
